Sampling Strategies in Decisions from Experience

### 2022-05-23 manuscript.Rmd
Update Section 4: Methods + Choice Rates (round-wise integration model)

To substantiate the discussion and obtain the actual predictions of the models, computational implementations thereof were used to simulate the sampling and accumulation process for a set of choices between a safe prospect and a two-outcome (risky) prospect.
The computational models were written in `R` and are embodied in the `GitHub` repository accompanying this paper.
The simulation was conducted for varying combinations of $\psi$ and $\theta$.
For each of these combinations, the simulated data was modeled with a stochastic version of CPT, using Bayesian parameter estimation.
The simulated choices and estimated CPT parameters can be roughly understood as theory-implied data and data models, which remain to be tested against empirical data.

@@

A test set of $60$ choice problems was obtained by stratified sampling from an initial set of $10,000$ problems.
The stratification procedure was used to ensure that the test set contained choice problems with small-probability outcomes of different rank---i.e., small-probability outcomes that are either higher or lower than the EV of the risky prospect---as well as choice problems without small-probability outcomes.
The exact procedure was as follows:
For each of the $10.000$ problems in the initial set, three outcomes were randomly drawn from a uniform distribution over the interval $[0, 20]$, with the smallest and highest of these three outcomes being assigned to the risky prospect to omit dominance.^[I.e., when one prospect always produces a higher outcome.]
The probability for the lower outcome of the risky prospect, $p_x(x_{low})$, was drawn from a uniform distribution over the interval $(0, 1)$, with the probability of the higher outcome being set to $p_x(x_{high}) = 1-p_x(x_{low})$.
To obtain the $60$ choice problems of the test set, the initial set was divided into three subsets, where each subset contained either all problems with $p_x(x_{high}) \in (0,.2)$ or $p_x(x_{high}) \in [.2,.8]$ or $p_x(x_{high}) \in (.8,1)$.
From each of these subset, then, $20$ choice problems were randomly sampled.

@@

* *Threshold*: 
$\theta$ was varied in the interval $[1,5]$ in increments of 1.

* *Decision variable*: The decision variable of a prospect either counts the number of won comparisons---implying an absolute threshold---or how many more comparisons were won relative to the other prospect---implying a relative threshold.^[<!--Open footnote-->
The results are similar for absolute and relative thresholds. 
For simplification, only the results of the absolute thresholds are reported in the main text.
The results for relative thresholds are reported in the Appendix.
<!--Close footnote-->]

@@

\noindent
Since $\psi$ introduces some stochasticity into the sampling and accumulation process, each of the 100 resulting parameter combinations was used 100 times---resembling 100 independent agents---to solve the test set. 
In sum, then 100 (parameter combinations) $\times$ 100 (agents) $\times$ 60 (choice problems)  = 600,000 sampling and accumulation processes were simulated.
The respective dataset can be retrieved from the accompanying `GitHub repository`.

@@

The choice data for each parameter combination was separately modeled in CPT, using the value function of @tverskyAdvancesProspectTheory1992 and the weighting function of @goldsteinExpressionTheoryPreference1987.
To demonstrate that the distortions from linear probability weighting are not just due to differences between the latent objective probabilities and the relative frequencies with which the outcomes were actually sampled over the course of a sampling and accumulation process, the latter were supplied as probability information to the weighting function.
To accommodate for possible random choice behavior that may be exerted by a parameter combination, the CPT model was amended with the logit choice rule
$$
p(P'^{safe}) = \frac{1} {1 + e^{-\rho(V_{safe}^{\frac{1}{\alpha}}-V_{risky}^{\frac{1}{\alpha}})}}
\; ,
\tag{16}
$$
where $p(P'^{safe})$ is the probability that the safe prospect is chosen over the risky prospect, $V_{safe}$ and $V_{risky}$ are the valuations of the safe and risky prospect that result from the parameter estimates of the value and weighting function, and $\rho \geq 0$ is a sensitivity parameter.
$\rho$ governs how strongly the choice between prospects depends on the differences in the valuations of the prospects, i.e., the core CPT model.
Specifically, $\rho = 0$ implies no sensitivity, and thus a random choice.
In turn, for increasing values of $\rho$, the probability of choosing the prospect with the higher valuation increases.
Note further that the valuation of the prospects were rescaled by $\frac{1}{\alpha}$, as proposed by @stewartPsychologicalParametersHave2018, to guard against possible parameter interdependencies between $\alpha$ and $\rho$ [@krefeld-schwalbStructuralParameterInterdependencies2022; @scheibehenneUsingBayesianHierarchical2015].
In sum, then, the model has four free parameters: $\alpha, \gamma, \delta, \rho$.

Since for a given choice problem and parameter combination of the sampling and accumulation model there should be no systematic differences among the 100 iterations (synthetic agents), their data is combined for the estimation of the free parameters.
Such a grouped approach should reveal the main effects of a given parameter combination of the sampling and accumulation model on the resulting choices and ignore unsystematic differences or noise that may be caused by the stochasticity introduced by $\psi$.
To facilitate comparisons across parameter combinations of the sampling and accumulation model, uninformative prior distributions were used throughout for all four parameters.
Specifically, the prior distributions $\alpha \sim U(0,1)$, $\gamma \sim U(0,2)$, $\delta \sim U(0,10)$, and $\rho \sim U(0,5)$ were used.

To estimate the posterior distributions of the free parameters, Markov Chain Monte Carlo sampling as implemented in JAGS was used, where 20 chains of 40,000 samples each were run after a warm-up period of 1000 samples.
To reduce potential autocorrelation during the sampling process, only every 20th sample was kept (thinning), leaving a total of 40,000 samples across chains.
The minimum effective sample size was `r min(cpt$n.eff)`.
The potential scale reduction factor $\hat{R}$ [@gelmanInferenceIterativeSimulation1992] was $\leq$ `r round(max(cpt$Rhat), 4)` for all models and parameters, indicating good convergence.

@@

Overall, the results of the simulation match the predictions from Table\ \@ref(tab:predictions), however, with some exceptions concerning the estimates for the parameter $\delta$ of the weighting function. 
We start by inspecting how the length of the information-processing chain changes with varying combinations of $\psi$ and $\theta$.

### Simulated Length of the Information-Processing Chains

Figure\ \@ref(fig:chain-length) displays the median number of outcomes sampled over the course of a sampling and accumulation process for each parameter combination.
The model predicts a decrease in the overall number of sampled outcomes with increases in the switching probability $\psi$.
As can be seen from the change in the slopes of the lines in Figure\ \@ref(fig:chain-length), the magnitude of this decrease diminishes as the relative increase in $\psi$ diminishes.^[<!--Open footnote-->I.e., the ratio of high values of $\psi$ is smaller than the ratio of small values of $\psi$.<!--Close footnote-->]
Moreover, the overall number of sampled outcomes increases with thresholds at a relatively constant rate.
Consider also the dashed horizontal line in Figure\ \@ref(fig:chain-length), which displays the median number of 14 outcomes that @wulffMetaanalyticReviewTwo2018 found to be sampled in 10,712 trials involving a choice between a safe and a two-outcome prospect in the sampling paradigm.
Notably, the parameter combinations that closely resemble the round-wise strategy---i.e., high values of $\psi$ and $\theta$---or the summary strategy---i.e., low values of $\psi$ and $\theta$---lead to overall numbers of sampled outcomes that come close to the meta-analytic median.
Moreover, while parameter combinations that can be understood as falling in between the round-wise and the summary strategy also lead to overall numbers of sampled outcomes that approximate the meta-analytic median, extreme combinations produce numbers that are considerably lower or higher, and thus less plausible from an empirical stance.

@@

\begin{figure}[t]
\caption{Lengths of Information-Processing Chains}
\includegraphics[width=1\linewidth]{manuscript_files/chain-length} \label{fig:chain-length}
\raggedright \textit{Note.} Add note
\end{figure}

### Simulated Final Choice

Figures\ \@ref(fig:choice-rates-roundwise) and \@ref(fig:choice-rates-summary) display the rates of choices that did not maximize the latent EV (hereafter: false response rates) for each combination of $\psi$ and $\theta$.
To demonstrate that the deviations from EV maximization are not just due to sampling error, the rates of choices that did not maximize the sampled mean are displayed in gray. 
Rates are separated for presence and rank of small-probability outcomes.
Consider first the top panel showing the false response rates across all problems where the low-rank outcome of the risky prospect possesses a small probability:
Over the course of short stochastic processes, then, the low-rank outcome tends to be sampled less often than would be objectively warranted, leading to an inflation of the respective mean of the risky prospect.
Accordingly, the top panels show that the rates of false choices of the risky prospect increase with $\psi$, but the rates of false safe choices remain generally low.
More specifically, the higher the values of $\psi$, the more severe should the inflation of the mean of the risky prospect be. 
As a consequence, if the risky prospect possesses the smaller EV, an increase in $\psi$ increases the risk to deviate from EV maximization---taking the form of an as-if underweighting of rare outcomes---by falsely choosing the risky prospect on the basis of a mean comparison.
In turn, if the risky prospect possesses the larger EV, the inflation of their means decreases the risk to falsely choose the safe prospect.
Note from the middle panel that the pattern of false response rates reverses for problems where the high-rank outcome of the risky prospect possesses a small probability.
In this case, then, the mean of the risky prospect tends to be more deflated the higher the value of $\psi$, which increases the risk of falsely choosing the safe prospect.
From the reduced false response rates in the bottom panel, we see that the inflation and deflation of the means for high values of $\psi$ is significantly attenuated when a prospect possesses no small-probability outcome.

@@

### Simulation-Implied Cumulative Prospect Theory

@@

>>>>

To substantiate the above discussion and obtain the actual predictions of the models, computational implementations thereof were used to simulate the sampling and accumulation process for a set of choices between a safe prospect and a two-outcome (risky) prospect.
The computational models were written in `R` and are embodied in the `GitHub` repository accompanying this paper.
The simulation was conducted for varying combinations of the switching probability $\psi$ and threshold $\theta$.
For each of these combinations, the simulated data was modeled with a stochastic version of CPT, using Bayesian parameter estimation.
The simulated data and estimated CPT parameters can be understood as theory-implied data and data models which remain to be tested against empirical data.

@@

A test set of 60 choice problems was obtained by stratified sampling from an initial set of 10,000 problems.
The stratification procedure was used to ensure that the test set contained choice problems with small-probability outcomes of different rank---i.e., small-probability outcomes that are either higher or lower than the EV of the risky prospect---as well as choice problems without small-probability outcomes.
The exact procedure was as follows:
For each of the 10,000 problems in the initial set, three outcomes were randomly drawn from a uniform distribution over the interval $[0, 20]$, with the smallest and highest of these three outcomes being assigned to the risky prospect to omit dominance.^[I.e., when one prospect always produces a higher outcome.]
The probability for the lower outcome of the risky prospect, $p_x(x_{low})$, was drawn from a uniform distribution over the interval $(0, 1)$, with the probability of the higher outcome being set to $p_x(x_{high}) = 1-p_x(x_{low})$.
To obtain the 60 choice problems of the test set, the initial set was divided into three subsets, where each subset contained either all problems with $p_x(x_{high}) \in (0,.2)$ or $p_x(x_{high}) \in [.2,.8]$ or $p_x(x_{high}) \in (.8,1)$.
From each of these subset, then, $20$ choice problems were randomly sampled.

@@

* *Threshold*: 
Round-wise integration model: $\theta$ was varied in the interval $[1,5]$ in increments of 1. Summary integration model: $\theta$ was varied in the interval $[15, 75]$ in increments of 15.

* *Decision variable*: Round-wise integration model: The decision variable of a prospect either counts the number of won comparisons---implying an absolute threshold---or how many more or less comparisons were won relative to the other prospect---implying a relative threshold. Summary integration model: The decision variable of a prospect is either the cumulative sum across all sampled outcomes or the difference in cumulative sums, implying absolute and relative thresholds.^[<!--Open footnote-->
For both models, the results are similar for absolute and relative thresholds.
For simplification, only the results of the absolute thresholds are reported in the main text.
The results for relative thresholds are reported in the Appendix.
<!--Close footnote-->]

@@

\noindent
Since the switching probability $\psi$ introduces some stochasticity into the sampling and accumulation process, each of the 100 resulting parameter combinations was used 100 times---resembling 100 independent agents---to solve the test set. 
In sum, then 2 (models) $\times$ 100 (parameter combinations) $\times$ 100 (agents) $\times$ 60 (choice problems)  = 1,200,000 sampling and accumulation processes were simulated.
The respective datasets can be both recreated or directly retrieved from the accompanying `GitHub` repository.

@@

The choice data implied by each integration model and parameter combination was separately modeled in CPT, using the value function of @tverskyAdvancesProspectTheory1992 and the weighting function of @goldsteinExpressionTheoryPreference1987.
Again, to demonstrate that the distortions from linear probability weighting are not just due to sampling error, the relative frequencies with which the outcomes were actually sampled over the course of a sampling and accumulation process were supplied as probability information to the weighting function.
To accommodate for possible random choice behavior that may be exerted by a parameter combination, the CPT model was amended with the logit choice rule
$$
p(P^{safe}) = \frac{1} {1 + e^{-\rho(V_{safe}^{\frac{1}{\alpha}}-V_{risky}^{\frac{1}{\alpha}})}}
\; ,
\tag{12}
$$
where $p(P^{safe})$ is the probability that the safe prospect is chosen over the risky prospect, $V_{safe}$ and $V_{risky}$ are the valuations of the safe and risky prospect that result from the parameter estimates of the value and weighting function, and $\rho \geq 0$ is a sensitivity parameter.
$\rho$ governs how strongly the choices between prospects depend on the differences in the valuations of the prospects, i.e., the core CPT model.
Specifically, $\rho = 0$ implies no sensitivity, and thus a completely random choice.
In turn, for increasing values of $\rho$, the probability of choosing the prospect with the higher valuation increases.
Note further that the valuation of the prospects were rescaled by $\frac{1}{\alpha}$, as proposed by @stewartPsychologicalParametersHave2018, to guard against possible parameter interdependencies between $\alpha$ and $\rho$ [@krefeld-schwalbStructuralParameterInterdependencies2022; @scheibehenneUsingBayesianHierarchical2015].
In sum, then, the model has four free parameters: $\alpha, \gamma, \delta, \rho$.

Since for a given choice problem and parameter combination of the sampling and accumulation model there should be no systematic differences among the 100 iterations (synthetic agents), their data is combined for the estimation of the free parameters.
Such a grouped approach should reveal the main effects of a given parameter combination of the sampling and accumulation model on the resulting choices and ignore unsystematic differences or noise that may be caused by the stochasticity introduced by the switching probability $\psi$.^[<!--Open footnote-->Note that this is quite a hard assumption that works for this simulation but will be hardly met for empirical data, where heterogeneity among people is common. If heterogeneity is assumed, hierarchical implementation of CPT may be used [cf. @nilssonHierarchicalBayesianParameter2011] <!--Close footnote-->.]
To facilitate comparisons across parameter combinations of the sampling and accumulation models, uninformative prior distributions were used throughout for all four parameters.
Specifically, the prior distributions $\alpha \sim U(0,1)$, $\gamma \sim U(0,2)$, $\delta \sim U(0,10)$, and $\rho \sim U(0,5)$ were used.

```{r convergence-statistics, include=FALSE}
effsize <- min(cpt$n.eff)
rhat <- round(max(cpt$Rhat), 4)
```

To estimate the posterior distributions of the free parameters, Markov Chain Monte Carlo sampling as implemented in `JAGS` was used, where 20 chains of 40,000 samples each were run after a warm-up period of 1000 samples.
To reduce potential autocorrelation during the sampling process, only every 20th sample was kept (thinning), leaving a total of 40,000 samples across chains.
The minimum effective sample size was `r format(effsize, big.mark=",", trim=TRUE)`.
The potential scale reduction factor $\hat{R}$ [@gelmanInferenceIterativeSimulation1992] was $\leq$ `r rhat` for all models and parameters, indicating good convergence.

@@

Below, the round-wise and the summary integration model are discussed separately.
However, except for the choice rates, displayed in Figures\ \@ref(fig:choice-rates-roundwise) and \@ref(fig:choice-rates-summary), the results of both models are displayed in the same figures using a green and blue color-coding for the round-wise and the summary integration model, respectively.

Overall, the results of the simulations (model predictions) match the expectations formulated in Table\  \@ref(tab:expectations-roundwise), however, with some exceptions concerning the estimates for the parameter $\delta$ of the weighting function.
First, it is inspected how the lengths of the sampling and accumulation processes---measured in the number of sampled outcomes---changes with varying combinations of the switching probability $\psi$ and threshold $\theta$.

#### Number of Sampled Outcomes

Figure\ \@ref(fig:chain-length) displays the number of outcomes sampled over the course of a sampling and accumulation process for each parameter combination.
The model predicts a decrease in the overall number of sampled outcomes with increases in the switching probability $\psi$.
As can be seen from the changes in the slopes of the gray lines of the smoothed functions, the magnitude of this decrease diminishes as the relative increase in $\psi$ diminishes.
The overall number of sampled outcomes increases with thresholds at a relatively constant rate.
Consider also the dashed horizontal lines, which display the median number of 14 outcomes that @wulffMetaanalyticReviewTwo2018 found to be sampled in 10,712 trials involving a choice between a safe and a two-outcome prospect in the sampling paradigm.
The parameter combinations that closely resemble the links between sampling and decision strategies proposed by Hills and Hertwig [-@hillsInformationSearchDecisions2010]---i.e, high values of $\psi$ and $\theta$ for piecewise sampling and round-wise decisions vs. low values of $\psi$ and $\theta = 1$ for comprehensive sampling and summary decisions)---led to overall numbers of sampled outcomes that come close to the meta-analytic median found by @wulffMetaanalyticReviewTwo2018.
Moreover, while parameter combinations that can be understood as falling in between the originally proposed links---i.e., values in the mid-range for both $\psi$ and $\theta$---also led to overall numbers of sampled outcomes that approximate the meta-analytic median, extreme combinations of both models produce numbers that are considerably lower or higher, and thus less plausible from an empirical stance.

@@

\begin{figure}[t!]
\caption{Lengths of the Sampling and Accumulation Processes}
\includegraphics[width=1\linewidth]{manuscript_files/chain-length} \label{fig:chain-length}
\raggedright \textit{Note.} Each point represents a single sampling and accumulation process (choice trial) and is the overall number of outcomes sampled over the course of this process. The gray lines are estimates of \textit{ggplot2}'s \texttt{stat\_smooth()} function. All gray-dashed horizontal lines represent the meta-analytic median, med = 14, for choices between a safe and a risky prospect, found by \protect\hyperlink{ref-wulffMetaanalyticReviewTwo2018}{Wulff et al., 2018}. Note that for better resolution the scales on the y-axis differ for different levels of \(\theta\).
\end{figure}

#### Final Choice

For each combination of the switching probability $\psi$ and the threshold $\theta$, Figure\ \@ref(fig:choice-rates-roundwise) displays the rates of choices that did not maximize the sampled mean (i.e., the mean across all outcomes sampled over the course of a sampling and accumulation process, hereafter: false response rates).
Rates are separated for presence and rank of small-probability outcomes.
Consider first the top panel showing the false response rates across all problems where the low-rank outcome of the risky prospect possesses a small probability:
For high switching probabilities $\psi$, the low-rank outcome was expected to be sampled less often than would be objectively warranted *within each* comparison round, leading to an inflation of the respective round means of the risky prospect.
Accordingly, the top panels show that for each threshold $\theta$ the rates of false choices of the risky prospect increase with the switching probability $\psi$, thereby causing deviation from sampled mean maximization that take the form of an as-if underweighting of rare outcomes.
In turn, rates of false safe choices remain generally low.
To demonstrate how the false response rates would have turned out given the normative solutions of EV maximization---i.e., when sampling error is not controlled for---, the rates of choices that did not maximize the latent EV are displayed in gray. 
Here, a similar but more severe pattern can be found, showing that sampling error exercises an influence beyond the effects implied by the interplay of the sampling strategy and the round-wise integration/decision strategy.
Note from the middle panel that the pattern of false response rates reverses for problems where the high-rank outcome of the risky prospect possesses a small probability.
In this case, then, the means of the risky prospect within each comparison round tend to be more deflated the higher the switching probability $\psi$, which increases the risk of falsely choosing the safe prospect.
Again, the resulting choice patterns resemble an as-if underweighting of rare outcomes.
Importantly, note from the reduced false response rates in the bottom panel that the inflation and deflation of the means for high switching probabilities must be significantly attenuated when a prospect possesses no small-probability outcome.
In sum, then, the simulated choices show that the potential interplay between sampling strategies and a round-wise integration/decision strategy can lead to distinct choice patterns and as-if weighting patterns in DFE, which, however, strongly depend on the structure of the choice problems.

@@

#### CPT Parameters

@@

#### Number of Sampled Outcomes

#### Final Choice

#### CPT Parameters
